---
title: "S&DS 363 Homework 4"
author: "Evan Collins, Kelly Farley, Ken Stier"
date: "30 March 2021"
output:
  html_document:
   toc: yes
   toc_float:
     collapsed: no
  pdf_document:
    latex_engine : xelatex
---

```{r, echo=F, message=F, warning=F}
# clear env
rm(list=ls())

# Load packages
library(car)
library(tidyverse)
library(MASS)
library(biotools)
library(DiscriMiner)
library(klaR)
library(vegan)

#source("http://www.reuningscherer.net/Multivariate/R/CSQPlot.r.txt")
#The file wouldn't knit because the source link timed out, so I just copied the contents to run natively:
CSQPlot<-function(vars,label="Chi-Square Quantile Plot"){
  #usually, vars is xxx$residuals or data from one group and label is for plot
  x<-cov(scale(vars),use="pairwise.complete.obs")
  squares<-sort(diag(as.matrix(scale(vars))%*%solve(x)%*%as.matrix(t(scale(vars)))))
  quantiles<-quantile(squares)
  hspr<-quantiles[4]-quantiles[2]
  cumprob<-c(1:length(vars[,1]))/length(vars[,1])-1/(2*length(vars[,1]))
  degf<-dim(x)[1]
  quants<-qchisq(cumprob,df=degf)
  gval<-(quants**(-1+degf/2))/(exp(quants/2)*gamma(degf/2)*(sqrt(2)**degf))
  scale<-hspr / (qchisq(.75,degf)-qchisq(.25,degf))
  se<-(scale/gval)*sqrt(cumprob*(1-cumprob)/length(squares))
  lower<-quants-2*se
  upper<-quants+2*se
  
  plot(quants,squares,col='red',pch=19,cex=1.2,xlab="Chi-Square Quantiles",
       ylab="Squared MH Distance",main=paste("Chi-Square Quantiles for",label),ylim=range(upper,lower, squares) , xlim=range(c(0,quants)))
  lines(c(0,100),c(0,100),col=1)
  lines(quants,upper,col="blue",lty=2,lwd=2)
  lines(quants,lower,col="blue",lty=2,lwd=2)
  legend("topleft",c("Data","95% Conf Limits"),lty=c(0,2),col=c("red","blue"),lwd=c(2,2),
         pch=c(19,NA))
}
```

# Contributors

Evan Collins (evan.collins@yale.edu)

Kelly Farley (kelly.farley@yale.edu)

Ken Stier (ken.stier@yale.edu)

# The Dataset

*Raw dataset:*
COVID-19 infection and death statistics from U.S. counties (sourced from NYT), combined with economic, education, and population data (sourced from various government agencies) and also survey responses about mask-wearing frequencies (sourced from NYT). 3141 complete observations on 19 metric variables and 6 categorical variables. To avoid any outliers due to population size differences between counties, all variables are scaled as a percentage of population. Variable descriptions can be found [here](http://evancollins.com/variable_descriptions.html).

*Data of relevance for this pset:*

Categorical Predictor 1 (rural_urban_code): The Rural-Urban Codes are numbered 1-9 according to descriptions provided by the USDA. We will regroup codes 1 through 9 as into three groups: (1) "Urban" for codes 1-3, (2) "Suburban" for codes 4-6, and (3) "Rural" for codes 7-9.

Categorical Predictor 2 (region): Each county is associated with a state name, which we will group into regions as defined by the U.S. Census Bureau: Northeast, Midwest, South, and West.

3 Continuous Response Variables: A NYT survey about masking behaviors asked people whether they wear a mask in public when they expect to be within 6 feet of another person and calculated the responses for each county for never, rarely, sometimes, frequently, and always mask. We choose to look at the extremes and will examine 3 continuous response variables: never mask, sometimes mask, and always mask.


```{r}
raw <- readr::read_csv("https://evancollins.com/covid_and_demographics.csv")

# create categorical variables: rural-urban code (3 levels), region (4 variables)
raw <- 
  raw %>%
    mutate(region = case_when(
      State_Name %in% c("Washington", "Oregon", "California", "Nevada", "Idaho", "Montana", "Utah", "Arizona", "Wyoming", "Colorado", "New Mexico", "Alaska", "Hawaii") ~ "West",
      State_Name %in% c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Wisconsin", "Illinois", "Michigan", "Indiana", "Ohio") ~ "Midwest",
      State_Name %in% c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Tennessee", "Kentucky", "Alabama", "Georgia", "Florida", "South Carolina", "North Carolina", "Virginia", "West Virginia", "District of Columbia", "Delaware", "Maryland") ~ "South",
      State_Name %in% c("Pennsylvania", "New Jersey", "Connecticut", "Rhode Island", "Massachusetts", "New Hampshire", "Vermont", "Maine", "New York") ~ "Northeast"),
      rural_urban_code = case_when(
        Rural_Urban_Code_2013 %in% c(1, 2, 3) ~ "Urban",
        Rural_Urban_Code_2013 %in% c(4, 5, 6) ~ "Suburban",
        Rural_Urban_Code_2013 %in% c(7, 8, 9) ~ "Rural")
      )
raw$rural_urban_code <- as.factor(raw$rural_urban_code) # Rural is reference
```

# 1: Interactions Plot

```{r}
interaction.plot(raw$rural_urban_code, raw$region, raw$Never_Wear_Mask_Survey,
  lwd = 3, col = c("red", "blue", "black", "green"), trace.label = "Region", 
  xlab = "Rural-Urban Setting", main = "Interaction Plot for Never Mask")

interaction.plot(raw$rural_urban_code, raw$region, raw$Sometimes_Wear_Mask_Survey,
  lwd = 3, col = c("red", "blue", "black", "green"), trace.label = "Region", 
  xlab = "Rural-Urban Setting", main = "Interaction Plot for Sometimes Mask")

interaction.plot(raw$rural_urban_code, raw$region, raw$Always_Wear_Mask_Survey,
  lwd = 3, col = c("red", "blue", "black", "green"), trace.label = "Region", 
  xlab = "Rural-Urban Setting", main = "Interaction Plot for Always Mask")
```

There does appear to be interaction between the rural-urban setting and the region, as evidenced by the non-parallel lines on the interaction plots for never masking, sometimes masking, and always masking. In particular, the West region seems to have behaviors that most contradict those of other regions, particularly in the Western suburbs.

# 2: Two-Way MANOVA

Univariate:

```{r}
Anova(lm(Never_Wear_Mask_Survey ~ region*rural_urban_code, data = raw), type = 3)
Anova(lm(Sometimes_Wear_Mask_Survey ~ region*rural_urban_code, data = raw), type = 3)
Anova(lm(Always_Wear_Mask_Survey ~ region*rural_urban_code, data = raw), type = 3)
```

Univariate:

The region, rural-urban code, and their interaction are significant in all 3 univariate models for masking behaviors, with p-values << alpha = 0.05. Based on coefficients, region is a more important predictor of sometimes masking, then always masking, then never masking; rural-urban code is a more important predictor of never masking, then sometimes masking, then always masking. but overall a less important predictor than region. Their interaction is most important in never masking, then always masking, then sometimes masking.


Multivariate:

Overall, there are differences between region and rural-urban codes (all multivariate statistics are significant). All of the multivariate tests suggest there is an interaction effect between region and rural-urban code.

```{r}
multimod <- lm(cbind(Never_Wear_Mask_Survey, Sometimes_Wear_Mask_Survey, Always_Wear_Mask_Survey) ~ region*rural_urban_code, data = raw)
summary(Anova(multimod), univariate=T)
```

# 3 Contrasts (Univariate and Multivariate)

Let's do the following comparisons:

- Multivariate - Rural vs. Suburban 

- Multivariate - Rural vs. Urban 

- Univariate - Rural vs. Urban

- Multivariate - Interaction between Rural and Urban between regions Northeast and South

- Univariate - Interaction between Rural and Urban between regions Northeast and South


First, let's make a variable `catcomb` that combines both categorical variables.

```{r}
raw$catcomb <- paste(raw$rural_urban_code, raw$region, sep = "")
table(raw$catcomb)
```


```{r}
options(contrasts = c("contr.treatment", "contr.poly"))

# Make catcomb a factor
raw$catcomb <- as.factor(raw$catcomb) # RuralMidwest is reference level

# Multivariate - Fit one way MANOVA model
multimod2 <- lm(cbind(Never_Wear_Mask_Survey, Sometimes_Wear_Mask_Survey, Always_Wear_Mask_Survey) ~ catcomb, data = raw)
# Univariate - Fit one way ANOVA model just for Never_Wear_Mask_Survey
multimodNever <- lm(Never_Wear_Mask_Survey ~ catcomb, data = raw)

contrasts(raw$catcomb)
levels(raw$catcomb)
```

```{r}
# Get multivariate contrast for Rural vs. Suburban
linearHypothesis(multimod2, "catcombRuralNortheast + catcombRuralSouth + catcombRuralWest - catcombSuburbanMidwest - catcombSuburbanNortheast - catcombSuburbanSouth - catcombSuburbanWest = 0")
```

For the multivariate tests between Rural and Suburban shown above, we can see that the Pillai, Wilks, Hotelling-Lawley, and Roy values all have p=1.1358e-10 < 0.05. Thus, we reject the null hypothesis and conclude Rural and Suburban are significantly different.

```{r}
# Get multivariate contrast for Rural vs. Urban
linearHypothesis(multimod2, "catcombRuralNortheast + catcombRuralSouth + catcombRuralWest - catcombUrbanMidwest - catcombUrbanNortheast - catcombUrbanSouth - catcombUrbanWest = 0")
```

For the multivariate tests between Rural and Urban shown above, we can see that the Pillai, Wilks, Hotelling-Lawley, and Roy values all have p=2.22e-16 < 0.05. Thus, we reject the null hypothesis and conclude Rural and Urban are significantly different.

```{r}
# Get univariate contrast for Rural vs. Urban
linearHypothesis(multimodNever, "catcombRuralNortheast + catcombRuralSouth + catcombRuralWest - catcombUrbanMidwest - catcombUrbanNortheast - catcombUrbanSouth - catcombUrbanWest = 0")
```

For the univariate F-test between Rural and Urban shown above, we can see that the p<2.2e-16 < 0.05. Thus, we reject the null hypothesis and conclude Rural and Urban are significantly different for `Never_Wear_Mask_Survey`.

```{r}
#Get multivariate contrast for Northeast,South and Rural,Urban interaction
linearHypothesis(multimod2, "catcombRuralNortheast - catcombUrbanNortheast - catcombRuralSouth + catcombUrbanSouth = 0") 
```

For the multivariate test above evaluating the interaction between Rural and Urban between regions Northeast and South, the difference is not shown to be significantly different, as the p value (0.27918) of the multivariate tests is greater than 0.05.


```{r}
#Get univariate contrast for Northeast,South and Rural,Urban interaction
linearHypothesis(multimodNever, "catcombRuralNortheast - catcombUrbanNortheast - catcombRuralSouth + catcombUrbanSouth = 0")
```

For the univariate test above evaluating the interaction between Rural and Urban between regions Northeast and South, the difference in `Never_Mask_Survey` is not shown to be significantly different, as the p value (0.7188) is greater than 0.05.



# 4 Multiple-Response Linear Model

Let's add two other continuous variables as covariates to the model and fit as a multiple-response linear model. We will include `` and `Percent_Adults_Bachelors_or_Higher` as covariates.

Let's first plot the relationships between the covariates and the three response variables. 

```{r, message=F, warning=F}
names(raw)
# For Median_Household_Income_2019
ggplot(raw, aes(x=Never_Wear_Mask_Survey, y=Median_Household_Income_2019)) + geom_smooth(method = lm, color = "green") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Never_Wear_Mask_Survey vs. Median Household Income")

ggplot(raw, aes(x=Sometimes_Wear_Mask_Survey, y=Median_Household_Income_2019)) + geom_smooth(method = lm, color = "green") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Sometimes_Wear_Mask_Survey vs. Median Household Income")

ggplot(raw, aes(x=Always_Wear_Mask_Survey, y=Median_Household_Income_2019)) + geom_smooth(method = lm, color = "green") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Always_Wear_Mask_Survey vs. Median Household Income")

# For Percent_Adults_Bachelors_or_Higher
ggplot(raw, aes(x=Never_Wear_Mask_Survey, y=Percent_Adults_Bachelors_or_Higher)) + geom_smooth(method = lm, color = "blue") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Never_Wear_Mask_Survey vs. % Adults with Bachelor's or Higher")

ggplot(raw, aes(x=Sometimes_Wear_Mask_Survey, y=Percent_Adults_Bachelors_or_Higher)) + geom_smooth(method = lm, color = "blue") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Sometimes_Wear_Mask_Survey vs. % Adults with Bachelor's or Higher")

ggplot(raw, aes(x=Always_Wear_Mask_Survey, y=Percent_Adults_Bachelors_or_Higher)) + geom_smooth(method = lm, color = "blue") + geom_point(color = "red", cex=0.5, alpha=0.3) + labs(title="Always_Wear_Mask_Survey vs. % Adults with Bachelor's or Higher")
```


```{r}
options(contrasts = c("contr.sum", "contr.poly"))

multimod3 <- lm(cbind(Never_Wear_Mask_Survey, Sometimes_Wear_Mask_Survey, Always_Wear_Mask_Survey) ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)

#Multivariate results and univariate results with with type 3 Sum of squares
summary(Anova(multimod3, type = 3), univariate = T)
```

In the above output chunk labeled "Term: region", we can see region is a significant (p<2.22e-16) multivariate predictor.

In the above output chunk labeled "Term: rural_urban_code", we can see rural-urban code is a significant (p<2.22e-16) multivariate predictor.

In the above output chunk labeled "Term: Median_Household_Income_2019", we can see median household income is a significant (p=0.00081009) multivariate predictor.

In the above output chunk labeled "Term: Percent_Adults_Bachelors_or_Higher", we can see median household income is a significant (p<2.22e-16) multivariate predictor.

In the above output chunk labeled "Term: region:rural_urban_code", we can see the interaction between rural and rural-urban code is a significant (p<2.22e-16) multivariate predictor.

In the bottom chunk labeled "Type III Sums of Squares", for each response variables, we can see the type III sum of squares, type III F-tests, and associated p values. From these univariate results, we can see that region is significant for each of the three response variables (Never_Wear_Mask_Survey, Sometimes_Wear_Mask_Survey, Always_Wear_Mask_Survey). Moreover, rural-urban code is significant for each of the three response variables. Median household income is significant just for Never_Wear_Mask_Survey. Percent of adults with a bachelor's degree of higher is significant for all three response variables. And the interaction between region and rural-urban code is significant for each of the three response variables.

# 5 Chi-Square Quantile Plots

Looking at the CSQ plot for the existing model...

```{r}
CSQPlot(multimod3$residuals)
```

Ooh, I don't love the look of that chi-square quantile plot. Let's take a closer look at the linear model features.

```{r}
modnever <- lm(Never_Wear_Mask_Survey ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modsometimes <- lm(Sometimes_Wear_Mask_Survey ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modalways <- lm(Always_Wear_Mask_Survey ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
plot(modnever, which = c(1,2), pch=19)
plot(modsometimes, which = c(1,2), pch=19)
plot(modalways, which = c(1,2), pch=19)
```

That could do it. There's considerable heteroskedasticity in the data. It looks like we're going to have to try a boxcox transformation.

```{r}
bcnever <- MASS::boxcox(Never_Wear_Mask_Survey+1/1000000000 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw, optimize=TRUE)
lambdanever <- bcnever$x[which.max(bcnever$y)]
bcsometimes <- MASS::boxcox(Sometimes_Wear_Mask_Survey+1/1000000000 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw, optimize=TRUE)
lambdasometimes <- bcsometimes$x[which.max(bcsometimes$y)]
bcalways <- MASS::boxcox(Always_Wear_Mask_Survey+1/1000000000 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw, optimize=TRUE)
lambdaalways <- bcalways$x[which.max(bcalways$y)]
```

We have our lambdas locked and loaded. Note, we added a minuscule amount to the survey values. That's just because the boxcox function requires positive values, and there were some Aileens in the mix.

```{r}
raw$newNever <- (raw$Never_Wear_Mask_Survey)^lambdanever
raw$newSometimes <- (raw$Sometimes_Wear_Mask_Survey)^lambdasometimes
raw$newAlways <- (raw$Always_Wear_Mask_Survey)^lambdaalways

modnever2 <- lm(newNever ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modsometimes2 <- lm(newSometimes ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modalways2 <- lm(newAlways ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
plot(modnever2, which = c(1,2), pch=19)
plot(modsometimes2, which = c(1,2), pch=19)
plot(modalways2, which = c(1,2), pch=19)
```

After transformation, the heteroskedasticity looks to be fairly diminished. Good job, boxcox! Let's check out the chi square quantile plot for our linear model.

```{r}
multimod4 <- lm(cbind(newNever, newSometimes, newAlways) ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)

summary(Anova(multimod4, type = 3), univariate = T)
CSQPlot(multimod4$residuals)
```

That looks a lot better, especially at the low values. At higher values, the points start to stray outside the 95% CI. Perhaps boxcox isn't really the way to go here. At JDRS's recommendation, let's go for Logit instead.

```{r}
raw$newNever2 <- logit(raw$Never_Wear_Mask_Survey)
raw$newSometimes2 <- logit(raw$Sometimes_Wear_Mask_Survey)
raw$newAlways2 <- logit(raw$Always_Wear_Mask_Survey)

modnever3 <- lm(newNever2 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modsometimes3 <- lm(newSometimes2 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
modalways3 <- lm(newAlways2 ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)
plot(modnever3, which = c(1,2), pch=19)
plot(modsometimes3, which = c(1,2), pch=19)
plot(modalways3, which = c(1,2), pch=19)

multimod5 <- lm(cbind(newNever2, newSometimes2, newAlways2) ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)

summary(Anova(multimod5, type = 3), univariate = T)
CSQPlot(multimod5$residuals)
```

Not terribly happy with how that came out. Let's draw some histograms to see why.

```{r}
hist(raw$Never_Wear_Mask_Survey)
hist(raw$Sometimes_Wear_Mask_Survey)
hist(raw$Always_Wear_Mask_Survey)

hist(logit(raw$Never_Wear_Mask_Survey))
hist(logit(raw$Sometimes_Wear_Mask_Survey))
hist(logit(raw$Always_Wear_Mask_Survey))
```

That adds up. Partly literally -- some of the issue is that the variables are necessarily interrelated and interdependent, in that the five mask survey response variables all fall on a single multidimensional curve by the nature of how they're determined. It's probably best to leave them untransformed and say that the data just have an unusual quality, per JDRS's recommendation, or alternatively to merge the variables into a single metric and add different response variables, which JDRS suggested but said is not necessarily warranted for the purposes of this pset. Before giving up on transformation, let's try applying logit to just the "never" variable, since it's the only one that is horribly skewed and butting up against 0.

```{r}
multimod6 <- lm(cbind(newNever2, Sometimes_Wear_Mask_Survey, Always_Wear_Mask_Survey) ~ region*rural_urban_code + Median_Household_Income_2019 + Percent_Adults_Bachelors_or_Higher, data = raw)

summary(Anova(multimod6, type = 3), univariate = T)
CSQPlot(multimod6$residuals)
```

Our best options appear to be either applying a logit transformation to the never-wears-mask values, or to just not transform at all. Given the nature of the data, no transformation is probably the way to go. At least, that's what JDRS advised, and it makes sense.

# 6 MRPP Test

Our data set is not Overwhelmingly Large™, so there is a reasonable expectation that MRPP will be able to provide a satisfactorily reliable p-value. Let's give it a go.

```{r, eval=FALSE}
(mrppout <- mrpp(raw[,c("Never_Wear_Mask_Survey", "Sometimes_Wear_Mask_Survey", "Always_Wear_Mask_Survey")], raw$rural_urban_code))
```

Now, RStudio's knitting function seems to have some difficulty processing MRPP. That's not much of a surprise; it takes my computer long enough to run as is, and we've had our share of issues with things running fine but not knitting properly. In any case, we had to disable evaluation of the function in order to knit, and the would-be output is copied below.

```
Dissimilarity index: euclidean 
Weights for groups:  n 

Class means and counts:

      Rural  Suburban Urban 
delta 0.2015 0.1966   0.1838
n     1077   899      1165  

Chance corrected within-group agreement A: 0.07795 
Based on observed delta 0.1935 and expected delta 0.2099 

Significance of delta: 0.001 
Permutation: free
Number of permutations: 999
```

And those results look pretty good! The p-value is 0.001 according to the above output, so MRPP tells us that the multivariate means of the groups are significantly different.