---
title: "S&DS 363 Homework 5"
author: "Evan Collins, Kelly Farley, Ken Stier"
date: "30 March 2021"
output:
  html_document:
   toc: yes
   toc_float:
     collapsed: no
  pdf_document:
    latex_engine : xelatex
---

```{r, echo=F, message=F, warning=F}
# clear env
rm(list=ls())

# Load packages
library(car)
library(tidyverse)
library(MASS)
library(biotools)
library(DiscriMiner)
library(klaR)
library(vegan)

#source("http://www.reuningscherer.net/Multivariate/R/CSQPlot.r.txt")
#The file wouldn't knit because the source link timed out, so I just copied the contents to run natively:
CSQPlot<-function(vars,label="Chi-Square Quantile Plot"){
  #usually, vars is xxx$residuals or data from one group and label is for plot
  x<-cov(scale(vars),use="pairwise.complete.obs")
  squares<-sort(diag(as.matrix(scale(vars))%*%solve(x)%*%as.matrix(t(scale(vars)))))
  quantiles<-quantile(squares)
  hspr<-quantiles[4]-quantiles[2]
  cumprob<-c(1:length(vars[,1]))/length(vars[,1])-1/(2*length(vars[,1]))
  degf<-dim(x)[1]
  quants<-qchisq(cumprob,df=degf)
  gval<-(quants**(-1+degf/2))/(exp(quants/2)*gamma(degf/2)*(sqrt(2)**degf))
  scale<-hspr / (qchisq(.75,degf)-qchisq(.25,degf))
  se<-(scale/gval)*sqrt(cumprob*(1-cumprob)/length(squares))
  lower<-quants-2*se
  upper<-quants+2*se
  
  plot(quants,squares,col='red',pch=19,cex=1.2,xlab="Chi-Square Quantiles",
       ylab="Squared MH Distance",main=paste("Chi-Square Quantiles for",label),ylim=range(upper,lower, squares) , xlim=range(c(0,quants)))
  lines(c(0,100),c(0,100),col=1)
  lines(quants,upper,col="blue",lty=2,lwd=2)
  lines(quants,lower,col="blue",lty=2,lwd=2)
  legend("topleft",c("Data","95% Conf Limits"),lty=c(0,2),col=c("red","blue"),lwd=c(2,2),
         pch=c(19,NA))
}
```

# Contributors

Evan Collins (evan.collins@yale.edu)

Kelly Farley (kelly.farley@yale.edu)

Ken Stier (ken.stier@yale.edu)

# The Dataset

*Raw dataset:*
COVID-19 infection and death statistics from U.S. counties (sourced from NYT), combined with economic, education, and population data (sourced from various government agencies) and also survey responses about mask-wearing frequencies (sourced from NYT). 3141 complete observations on 19 metric variables and 6 categorical variables. To avoid any outliers due to population size differences between counties, all variables are scaled as a percentage of population. Variable descriptions can be found [here](http://evancollins.com/variable_descriptions.html).

*Data of relevance for this pset:*

Categorical Predictor 1 (rural_urban_code): The Rural-Urban Codes are numbered 1-9 according to descriptions provided by the USDA. We will regroup codes 1 through 9 as into three groups: (1) "Urban" for codes 1-3, (2) "Suburban" for codes 4-6, and (3) "Rural" for codes 7-9.

Categorical Predictor 2 (region): Each county is associated with a state name, which we will group into regions as defined by the U.S. Census Bureau: Northeast, Midwest, South, and West.

3 Continuous Response Variables: A NYT survey about masking behaviors asked people whether they wear a mask in public when they expect to be within 6 feet of another person and calculated the responses for each county for never, rarely, sometimes, frequently, and always mask. We choose to look at the extremes and will examine 3 continuous response variables: never mask, sometimes mask, and always mask.


```{r}
raw <- readr::read_csv("https://evancollins.com/covid_and_demographics.csv")

# create categorical variables: rural-urban code (3 levels), region (4 variables)
raw <- 
  raw %>%
    mutate(region = case_when(
      State_Name %in% c("Washington", "Oregon", "California", "Nevada", "Idaho", "Montana", "Utah", "Arizona", "Wyoming", "Colorado", "New Mexico", "Alaska", "Hawaii") ~ "West",
      State_Name %in% c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", "Iowa", "Missouri", "Wisconsin", "Illinois", "Michigan", "Indiana", "Ohio") ~ "Midwest",
      State_Name %in% c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", "Tennessee", "Kentucky", "Alabama", "Georgia", "Florida", "South Carolina", "North Carolina", "Virginia", "West Virginia", "District of Columbia", "Delaware", "Maryland") ~ "South",
      State_Name %in% c("Pennsylvania", "New Jersey", "Connecticut", "Rhode Island", "Massachusetts", "New Hampshire", "Vermont", "Maine", "New York") ~ "Northeast"),
      rural_urban_code = case_when(
        Rural_Urban_Code_2013 %in% c(1, 2, 3) ~ "Urban",
        Rural_Urban_Code_2013 %in% c(4, 5, 6) ~ "Suburban",
        Rural_Urban_Code_2013 %in% c(7, 8, 9) ~ "Rural")
      )
raw$rural_urban_code <- as.factor(raw$rural_urban_code) # Rural is reference
```

# 1 Distance Metrics
Think about what distance metrics are appropriate for your data based on data type. Write a few sentences about this.  Also think about whether you should standardize and/or transform your data (comment as appropriate).

# 2 Hierarchical Cluster Analysis
Try various forms of hierarchical cluster analysis.  Try at least two different distance metrics and two agglomeration methods.  Produce dendrograms and comment on what you observe.

# 3 If possible, run the SAS macro or the R function to think about how many groups you want to retain (i.e. get the plots of cluster distance, R-squared, etc).  If you can’t run this, discuss how many groups you think are present.  

# 4 Run k-means clustering on your data.  Compare results to what you got in 3.)  Include a sum of squares vs. k (number of clusters) plot and comment on how many groups exist.

# 5 Comment on the number of groups that seem to be present based on what you find above.

# 6 Write a few sentences describing your final groups.